\vspace{-.6in}Annotation-Based Empirical Performance Tuning\\ Using Orio

### 2009-02-12 doc/papers/ipdps09/results-long.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

This section discusses the performance evaluations of the integrated
PLuTo-Orio system (explained earlier in Section~\ref{sec:integration})
on the multicore Intel machine by using a number of application
kernels that are non-trivial to optimize and parallelize. We compare
the performance of the code tuned by Orio with the base code and the
PLuTo-generated codes. The PLuTo code was obtained via running the
base code with PLuTo-0.3.0~\cite{pluto030} using \texttt{--tile} to
employ loop tiling for L1 cache,
\texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling. Generating a parallel
version of PLuTo code needs an additional \texttt{--parallel}
option. Note that we used PLuTo's default tile sizes and unroll
factors to produce all PLuTo codes. By default, PLuTo always sets the
L1 tile sizes to 32x32 (or 32x32x32) and L2 tile sizes to 256x256 (or
256x256x256) depending on the loop nest depth. The default unroll
factors are always 64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming. Furthermore, all codes were compiled with Intel C
Compiler using \texttt{-O3} optimization flag to ensure
auto-vectorization and other ICC optimizations, and \texttt{-parallel}
(or \texttt{-openmp}, when manual parallelization using OpenMP was
used) to enable code parallelization.

In the following sections, we refer to the ICC-optimized base code as
'Compiler-optimized', the PLuTo-generated code with L1 tiling and
unroll/jamming as 'PLuTo (L1 tiling)', and the PLuTo-generated code
with L1 and L2 tilings and unroll/jamming as 'PLuTo (L1+L2
tiling)'. Also, the code tuned by Orio is referred to as 'PLuTo+Orio'.

@@

The performance of sequential 2-D FDTD codes for $tmax=500$ and
$nx=ny$ is presented in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base
code optimized by ICC performs better than PLuTo for small problem
sizes since all input arrays fit in the L2 cache, thus tiling
performed by PLuTo will only introduce more overhead. As the array
sizes become larger, the lack of data reuse impairs the base code's
performance, whereas the PLuTo performance remains steady. Code tuning
using Orio generates two code variants: one for small problem sizes
and another for large problem sizes. When the input array sizes are
small, Orio discovers that applying PLuTo's polyhedral transformations
is not beneficial, and therefore, it employs only its syntactic
transformations on the original FDTD code. As to tuning for large
problem size, Orio exploits some of the PLuTo's code transformations
and enhances it further with its syntactic optimizations, resulting in
performance constantly higher than both the base and PLuTo codes. By
tuning the sequential code with Orio, the attainable performance
enhancement over PLuTo is up to 86\%.

Figure~\ref{fig:fdtd-2d-cookie-par} displays the multicore performance
obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
while ICC is unable to auto-parallelize the code, PLuTo detects the
existence of pipelined parallelism and then parallelizes the
code. Optimizing and tuning by Orio additionally improves the PLuTo
performance by up to 78\%. Moreover, it can be seen that due to memory
contention between the two quad-core Intel processors, the performance
speedup of both the PLuTo and Orio codes slightly decreases when the
number of used cores is greater than four.

@@

%%BN: commented out since this needs more explanation (no room in this version)
%It is to be noted that collecting performance results of the PLuTo 
%code with L1 and L2 tilings was not possible because of the large code 
%size of the generated loop nest, which led to an impractically huge 
%compile time.

@@

The sequential performance results shown in
Figure~\ref{fig:seidel-cookie-seq} are when $T=500$. As the figure
shows, applying PLuTo's polyhedral tiling on the original code always
delivers performance boosts, which range from 126\% to
142\%. Performing two-level tiling (for both L1 and L2 caches) by
PLuTo yields 13\% lower performance than only using L1 tiling; and
this is also reflected in the best sequential code found by Orio,
where one-level tiling (for L1 cache only) is always performed. The
sequential speedup obtained from tuning the PLuTo code with Orio is
significant, ranging from 51\% to 133\%.

@@

Parallel performance for $T=500$ and $N=4000$ can be viewed in
Figure~\ref{fig:seidel-cookie-par}. Similar earlier trend also exists
in these performance results: PLuTo parallelizes the code through its
polyhedral transformations (i.e., skewing and tiling), while the ICC
compiler alone does not parallelize the loop. Empirical tuning via
Orio creates up to 122\% speedups over the one-level tiled PLuTo code,
and up to 548\% speedups over the two-level tiled PLuTo code. Another
finding is that tiling for both L1 and L2 caches contributes to a poor
scalability because the number of tiles can be less than the number of
processors, resulting in underutilization of the available resources.

@@

The sequential performance results of LU Decomposition are presented
in Figure~\ref{fig:lu-cookie-seq}. Like 2-D FDTD, the ICC-optimized
code is more efficient than the PLuTo-generated codes when the input
arrays are small and fit in the L2 cache. However, due to better data
locality, the performance of the PLuTo-tiled codes exceeds the base
code as the input arrays get larger. Again, there are two different
codes generated by Orio: one for small problem sizes and one for large
problem sizes. And only the code tuned for large input arrays employs
PLuTo's polyhedral transformations, prior to applying Orio's syntactic
transformations. Unlike 3-D Gauss-Seidel, this time Orio chooses to
apply both L1 and L2 tilings when generating the best tuned
code. Compared to the two PLuTo codes, the Orio-tuned code yields
performance speedups in the range of 26\% to 277\%.

Parallel performance results shown in Figure~\ref{fig:lu-cookie-par}
were obtained for $N=4000$. From the results, we observe that ICC is
not able to parallelize the code, whereas PLuTo achieves higher
performance by exploitation of multicore parallelism. Furthermore,
both PLuTo-generated codes have comparable performance numbers, with a
slightly better scalability on the one-level tiled code. By the use of
Orio tuning, the performances of the PLuTo base codes were improved by
a factor of 51\% to 120\%.

>>>>

This section discusses the performance evaluations of the integrated
Pluto-Orio system (explained earlier in Section~\ref{sec:integration})
on the multicore Intel machine by using a number of application
kernels that are non-trivial to optimize and parallelize. We compare
the performance of the code tuned by Orio with the base code and the
Pluto-generated codes. The Pluto code was obtained via running the
base code with Pluto-0.3.0~\cite{pluto030} using \texttt{--tile} to
employ loop tiling for L1 cache,
\texttt{--l2tile} to employ loop tiling for L2 cache, and
\texttt{--unroll} to employ loop unrolling. Generating a parallel
version of Pluto code needs an additional \texttt{--parallel}
option. Note that we used Pluto's default tile sizes and unroll
factors to produce all Pluto codes. By default, Pluto always sets the
L1 tile sizes to 32x32 (or 32x32x32) and L2 tile sizes to 256x256 (or
256x256x256) depending on the loop nest depth. The default unroll
factors are always 64 for 1-D unrolling and 8x8 for 2-D
unroll/jamming. Furthermore, all codes were compiled with Intel C
Compiler using \texttt{-O3} optimization flag to ensure
auto-vectorization and other ICC optimizations, and \texttt{-parallel}
(or \texttt{-openmp}, when manual parallelization using OpenMP was
used) to enable code parallelization.

In the following sections, we refer to the ICC-optimized base code as
'Compiler-optimized', the Pluto-generated code with L1 tiling and
unroll/jamming as 'Pluto (L1 tiling)', and the Pluto-generated code
with L1 and L2 tilings and unroll/jamming as 'Pluto (L1+L2
tiling)'. Also, the code tuned by Orio is referred to as 'Pluto+Orio'.

@@

The performance of sequential 2-D FDTD codes for $tmax=500$ and
$nx=ny$ is presented in Figure~\ref{fig:fdtd-2d-cookie-seq}. The base
code optimized by ICC performs better than Pluto for small problem
sizes since all input arrays fit in the L2 cache, thus tiling
performed by Pluto will only introduce more overhead. As the array
sizes become larger, the lack of data reuse impairs the base code's
performance, whereas the Pluto performance remains steady. Code tuning
using Orio generates two code variants: one for small problem sizes
and another for large problem sizes. When the input array sizes are
small, Orio discovers that applying Pluto's polyhedral transformations
is not beneficial, and therefore, it employs only its syntactic
transformations on the original FDTD code. As to tuning for large
problem size, Orio exploits some of the Pluto's code transformations
and enhances it further with its syntactic optimizations, resulting in
performance constantly higher than both the base and Pluto codes. By
tuning the sequential code with Orio, the attainable performance
enhancement over Pluto is up to 86\%.

Figure~\ref{fig:fdtd-2d-cookie-par} displays the multicore performance
obtained for $tmax=500$ and $nx=ny=2000$. The results indicate that
while ICC is unable to auto-parallelize the code, Pluto detects the
existence of pipelined parallelism and then parallelizes the
code. Optimizing and tuning by Orio additionally improves the Pluto
performance by up to 78\%. Moreover, it can be seen that due to memory
contention between the two quad-core Intel processors, the performance
speedup of both the Pluto and Orio codes slightly decreases when the
number of used cores is greater than four.

@@

%%BN: commented out since this needs more explanation (no room in this version)
%It is to be noted that collecting performance results of the Pluto 
%code with L1 and L2 tilings was not possible because of the large code 
%size of the generated loop nest, which led to an impractically huge 
%compile time.

@@

The sequential performance results shown in
Figure~\ref{fig:seidel-cookie-seq} are when $T=500$. As the figure
shows, applying Pluto's polyhedral tiling on the original code always
delivers performance boosts, which range from 126\% to
142\%. Performing two-level tiling (for both L1 and L2 caches) by
Pluto yields 13\% lower performance than only using L1 tiling; and
this is also reflected in the best sequential code found by Orio,
where one-level tiling (for L1 cache only) is always performed. The
sequential speedup obtained from tuning the Pluto code with Orio is
significant, ranging from 51\% to 133\%.

@@

Parallel performance for $T=500$ and $N=4000$ can be viewed in
Figure~\ref{fig:seidel-cookie-par}. Similar earlier trend also exists
in these performance results: Pluto parallelizes the code through its
polyhedral transformations (i.e., skewing and tiling), while the ICC
compiler alone does not parallelize the loop. Empirical tuning via
Orio creates up to 122\% speedups over the one-level tiled Pluto code,
and up to 548\% speedups over the two-level tiled Pluto code. Another
finding is that tiling for both L1 and L2 caches contributes to a poor
scalability because the number of tiles can be less than the number of
processors, resulting in underutilization of the available resources.

@@

The sequential performance results of LU Decomposition are presented
in Figure~\ref{fig:lu-cookie-seq}. Like 2-D FDTD, the ICC-optimized
code is more efficient than the Pluto-generated codes when the input
arrays are small and fit in the L2 cache. However, due to better data
locality, the performance of the Pluto-tiled codes exceeds the base
code as the input arrays get larger. Again, there are two different
codes generated by Orio: one for small problem sizes and one for large
problem sizes. And only the code tuned for large input arrays employs
Pluto's polyhedral transformations, prior to applying Orio's syntactic
transformations. Unlike 3-D Gauss-Seidel, this time Orio chooses to
apply both L1 and L2 tilings when generating the best tuned
code. Compared to the two Pluto codes, the Orio-tuned code yields
performance speedups in the range of 26\% to 277\%.

Parallel performance results shown in Figure~\ref{fig:lu-cookie-par}
were obtained for $N=4000$. From the results, we observe that ICC is
not able to parallelize the code, whereas Pluto achieves higher
performance by exploitation of multicore parallelism. Furthermore,
both Pluto-generated codes have comparable performance numbers, with a
slightly better scalability on the one-level tiled code. By the use of
Orio tuning, the performances of the Pluto base codes were improved by
a factor of 51\% to 120\%.
